---
title: "Project 1"
author: "Ethan Marburger, David Pressley"
format: html
editor: visual
render-on-save: true
---

## Project 1

### Introduction

The purpose of Project One is to query, summarize, and display data using the census.gov API

### Workflow

The main workflow consists of

1.  Calling a function to get data from the API
2.  Checking the returned data meet criteria defined in the project specification
3.  Getting metadata for specified variables
4.  Mapping classification levels for factor variables
5.  Mapping time variables to the midpoint for times that have ranges and formatting in 24hour time format
6.  Returning a tibble of the API
7.  Calling a function that allows the user to specify multiple years of survey data when querying the API and returns results as a tibble
8.  Making our previous tibble have an additional class of "census"
9.  Calling a function that produces means and standard deviations for numeric variable(s) and counts for categorical variable(s) in a named list
10. Create boxplot of selected variables:
    - Box plot of Age (AGEP) by Sex
    - Box plot of Broadband Access (HISPEED) by Sex
    - Box plot of School Enrollment (SCH) by Age (AGEP) < 30
    - Box plot of Age < 30 vs. School Enrollment by SEX

### Implementation

#### Include required libraries

```{r}
library(readr)
library(tidycensus)
library(jsonlite)
library(httr)
library(stringr)
library(lubridate)
library(janitor)
library(tidyverse)

#store key
census_api_key("bdb1f6ff2e4982a1da64cd526532aa92dca5581c", install = TRUE, overwrite = TRUE)
readRenviron("~/.Renviron")
```

### Data Processing: Workflow Steps 1—6

We created a few helper functions to ensure separation of concerns and modularity of code. The implementation of these functions is detailed below

#### Function: `return_tibble`

-   returns data from an API call to census.gov as a tibble

```{r}
#return API query as tibble
return_tibble <- function (content)
{
  parsed_data <- as_tibble(fromJSON(rawToChar(content)))
  
  #set column names from first row
  col_names <- parsed_data[1,]
  parsed_data <- parsed_data[-1,]
  parsed_data <- setNames(parsed_data, col_names) 
  
  #save data to disk
  # write_csv(parsed_data, "raw_api.csv")
  return(parsed_data)  
}
```

#### Function: `get_var_metadata`
- dynamically gets variable metadata and returns a lookup table for specified variables in 
  function `get_PUMS`
- this function is currently of limited utility, but works for the current project. 
- the original idea was to create a metadata table from the project specification 
that could be used to lookup and drive the logic for data processing.
  - It needs work to be part of a consistent metadata driven approach, but works minimally 
    for this project

```{r}
#get metadata from specification for user provided variables
#metadata can be used for further checks, type conversion, and factor leveling
#get_var_metadata currently checks for required variables in a user provided
#function call, but no other checks
get_var_metadata <- function(var_list){
  spec_vars <- tibble(varname = c("PWGTP","AGEP","GASP","GRPIP","JWAP","JWDP",
                                  "JWMNP","FER","HHL","HISPEED","JWTRNS","SCH",
                                  "SCHL","SEX"),
                      type = c("num","num","num","num","chr","chr","num","chr"
                               ,"chr","chr","chr","chr","chr","chr"),
                      required = c(TRUE,TRUE,TRUE,TRUE,TRUE,TRUE,TRUE,TRUE,
                                   TRUE,TRUE,TRUE,TRUE,TRUE,TRUE),
                      dt = c(FALSE,FALSE,FALSE,FALSE,TRUE,TRUE,TRUE,FALSE,FALSE,FALSE,
                             FALSE,FALSE,FALSE,FALSE)
                      )
  
  #unlist user specified / API required vars to vector
  if (!"PWGTP" %in% var_list){
    var_list <- c(var_list, "PWGTP")
  }

  var_list <- unlist(str_split(var_list, ","))
  var_list <- str_trim(var_list)
  var_list <- var_list[var_list != "" & var_list != " "]


  #return required varnames as a vector to use in quality checks
  user_vars <- spec_vars |>
    filter(varname %in% var_list)

  #return tibble for varnames in user provided list. This is the lookup table to
  #be used for checking variable types and factor levels
  return(user_vars)
  }

```

#### Function: `check_vars`
-   checks `geography` and `year` against specified values

```{r}
check_vars <- function(user_vars, geography, year)
{
  year_range <- seq(2010,2022, by = 1)
  if (!geography %in% c("all","region","division","state")){
    stop(paste(geography), " is not an allowed value for the geography variable. Options are: all, region, division, state")
  } 
  if (!year %in% year_range){
    stop(paste(year), " is not an allowed value for the year variable. Options are between 2010 - 2022")
  } 
  else if (year == 2020){
    stop(paste(year), " is not available from census.gov")
  }
  else{
    return(TRUE)
  }
}
```

#### Function: `get_pums_factor_levels` and `set_factor_levels_types`

-   getter function gets factor levels and returns them as a list. Initial designs included this, but are not referenced in subsequent data processing steps. Kept as part of project for posterity
-   setter function checks, processes, and sets classification levels for factor variables.
    -   returns a tibble

```{r}
#get factor levels
get_pums_factor_levels <- function(var_names){
  print(var_names)
  #initialize list
  levels_list <- list()
  var_names <- unlist(strsplit(var_names,","))
  print(var_names)
  #get base URL
  for (var_name in var_names){
    url <- paste0("https://api.census.gov/data/2022/acs/acs1/pums/variables/",var_name,".json")
    print(url)
    response <- GET(url)
    if (status_code(response) == 200) {
      data <- fromJSON(content(response,"text"))
      levels <- data$values$item
      levels_list[[var_name]] <- levels
    }
  }
  return(levels_list)
}

#get factor levels and set types
set_factor_levels_types <- function(data_tibble, var_tibbs) {
  # Process character variables and set them as factors with levels
  chr_vars <- var_tibbs|>
    filter(type == "chr") |>
    pull(varname)
  
  # Loop through chr_vars and fetch levels for factor variables
  data_tibble <- data_tibble |>
    mutate(across(all_of(chr_vars), ~ {
      var_name <- cur_column()  # Get the current column name
      url <- paste0("https://api.census.gov/data/2022/acs/acs1/pums/variables/", var_name, ".json")

      response <- GET(url)
      if (status_code(response) == 200) {
        data <- fromJSON(content(response, "text", encoding = "UTF-8"))
        
        if (!is.null(data$values) && "item" %in% names(data$values)) {
          levels <- names(data$values$item)
          labels <- data$values$item
          return(factor(.x, levels = levels, labels = labels))
          
        } else {
          warning(paste("Levels for", var_name, "not found in API response"))
          return(.x)  # Return unchanged column
        }
      } else {
        warning(paste("Failed to fetch data for", var_name, ": HTTP status", status_code(response)))
        return(.x)  # Return unchanged column
      }
    }))

  # Process numeric variables and convert them to numeric
  num_vars <- var_tibbs |>
    filter(type == "num") |>
    pull(varname)

  data_tibble <- data_tibble |> 
    mutate(across(all_of(num_vars), as.numeric))

  return(data_tibble)
}


#Checks

# x <- get_pums_factor_levels("JWMNP")
# devdata_tibbs <- pums_data$api_data
# glimpse(devdata_tibbs)
# devvar_tibbs <- pums_data$api_metadata
# glimpse(devvar_tibbs)
# devx <- set_factor_levels_types(devdata_tibbs, devvar_tibbs)

# print(unique(devx$JWAP))
# print(unique(devx$SEX))
# print(unique(devx$FER))
```

#### Function: `extract_times`, `convert_to_24hr`, `calculate_midpoint`

-   these functions are dependent on one another. They convert factored time variables that contain a specific string pattern into a 24 hour time format representing the midpoint of the range
    -   Example: the factored string `3:00 p.m. to 3:04 p.m.` would be returned from this function as `15:02`

```{r}
extract_times <- function(jwap_value){
  # Handle N/A and missing cases
  if (str_detect(as.character(jwap_value), "N/A") | 
      is.na(jwap_value) |
      str_detect(as.character(jwap_value), "NA")) {
        return(list(first_24 = "NA", second_24 = "NA"))
  }
  
  # Split the string into parts (e.g., "9:40 p.m. to 9:44 p.m.")
  parts <- str_split(as.character(jwap_value), " ")[[1]]
  first_time <- parts[1]
  first_ampm <- str_replace_all(parts[2], "\\.", "")
  second_time <- parts[4]
  second_ampm <- str_replace_all(parts[5], "\\.", "")
  
  # Convert both times to 24-hour format
  first_24 <- convert_to_24_hour(first_time, first_ampm)
  second_24 <- convert_to_24_hour(second_time, second_ampm)
  
  return(list(first_24 = first_24, second_24 = second_24))
}

convert_to_24_hour <- function(time_str, ampm_str) {
  # Clean up the AM/PM string by removing periods
  full_time_str <- paste(time_str, ampm_str)  # Combine time and cleaned AM/PM
  time_parsed <- parse_date_time(full_time_str, "I:M p")  # Parse AM/PM time
  return(format(time_parsed, "%H:%M"))  # Convert to 24-hour format
  
}

calculate_midpoint <- function(time_parsed){
  second_24 <- as.POSIXct(time_parsed$second_24, format = "%H:%M")
  first_24 <- as.POSIXct(time_parsed$first_24, format = "%H:%M")
  
  #handle NA
  if(is.na(second_24) | is.na(first_24)) {
    return(NA)
  }
  
  midpoint <- first_24 + (second_24 - first_24)/2
  # print(format(midpoint, "%H:%M"))
  return(format(midpoint, "%H:%M"))
}


#Testing
# df <- pums_data$api_data |>
#   mutate(
#     time_parsed = map(JWAP, extract_times),  # Extract times into a list of parsed times
#     first_24 = map_chr(time_parsed, "first_24"),  # Extract first 24-hour time
#     second_24 = map_chr(time_parsed, "second_24"),  # Extract second 24-hour time
#     midpoint = map_chr(time_parsed, calculate_midpoint)  # Calculate midpoint between times
#   )
 # filtered_data <- pums_data$api_data |>
   # filter(str_detect(as.character(JWAP), "N/A") | is.na(JWAP) | str_detect(as.character(JWAP), "NA"))|>
   # filter(!(as.character(JWAP) %in% (c("N/A","NA", "", " ")))) |>
   # slice_head(n=100)
   
```

#### Function: `get_PUMS`

-   This function serves as the function to get, check, factor, and format variables according to project specifications

```{r}

get_PUMS <- function(geography, num_vars = NULL, chr_vars = NULL , key = sys.getenv("CENSUS_API_KEY"),
                             year = 2022, state = "05",county = NULL, zcta = NULL,
                             survey = "acs1", show_call = FALSE)
{
  #munge var lists
  if(is.null(num_vars)){
    num_vars <- c("PWGTP", "AGEP")
  }
  if(is.null(chr_vars)){
    chr_vars <- c("SEX")
  }

  user_vars <-c(num_vars, chr_vars)

  #get var metadata
  var_metadata <- get_var_metadata(user_vars)

  #get vars string for API call after returning metadata
  api_var_string <- paste(var_metadata$varname,collapse = ",")

  #check vars values in range
  var_check <- check_vars(user_vars,geography,year)
  # If check_vars returns a message (missing vars), stop the function
  if (!isTRUE(var_check)) {
    stop(var_check)  # Stops and prints the error message from check_vars
  }

  #get base URL
  baseURL <- paste("https://api.census.gov/data",
                   as.character(year), "acs",
                   survey, "pums", sep = "/")


  paramsURL <- paste0("?get=", api_var_string, "&for=", geography, ":", state, "&key=", key)
  apiURL <- (paste0(baseURL,paramsURL))

  #check
  print(apiURL)

  response <- httr::GET(apiURL)

  # Convert API response to a tibble |>
  #set factor levels and convert data types

  parsed_data <- return_tibble(response$content)
  jwmnptst <- return_tibble(response$content)
  # |>
  #   clean_names()
  # Set factor levels and types for the tibble

  tst <- set_factor_levels_types(parsed_data, var_metadata)
  
  parsed_data <- set_factor_levels_types(parsed_data, var_metadata)
  
  # df <- parsed_data$JWAP |>
  #     mutate(
  #       time_parsed = map(JWAP, extract_times),  # Extract times
  #       midpoint = map_chr(time_parsed, calculate_midpoint)  # Calculate midpoints
  #     )
  datetime_var_chk <- c("JWAP", "JWDP", "JWMNT")
  
  for (varname in datetime_var_chk){
    if (varname %in% names(parsed_data)){
      if (varname == "JWAP"){
        parsed_data <- parsed_data |>
          mutate(
            JWAP = as.character(JWAP),
            time_parsed = map(JWAP, extract_times),  # Extract times into a list of parsed times
            first_24 = map_chr(time_parsed, "first_24"),  # Extract first 24-hour time
            second_24 = map_chr(time_parsed, "second_24"),  # Extract second 24-hour time
            midpoint = map_chr(time_parsed, calculate_midpoint)) |>
          rename(jwap_t = midpoint) |>
          select(-first_24, -second_24 , -time_parsed)
      }
      else if (varname == "JWDP"){
        parsed_data <- parsed_data |>
          mutate(
            JWDP = as.character(JWDP),
            time_parsed = map(JWDP, extract_times),  # Extract times into a list of parsed times
            first_24 = map_chr(time_parsed, "first_24"),  # Extract first 24-hour time
            second_24 = map_chr(time_parsed, "second_24"),  # Extract second 24-hour time
            midpoint = map_chr(time_parsed, calculate_midpoint)) |>
          rename(jwdp_t = midpoint) |>
          select(-first_24, -second_24 , -time_parsed)
      }
      else if (varname == "JWMNP"){
        parsed_data <- parsed_data |>
          # filter(JWMNP == as.character("0")) |>
          mutate(
            jwmnp_c = as.character(JWMNP),
            jwmnp_n = suppressWarnings(as.numeric(jwmnp_c)),  # Suppress warnings and handle NA conversion
            jwmnp_mins = ifelse(is.na(jwmnp_n), NA, jwmnp_n)
          ) |>
          select(-jwmnp_c, -jwmnp_n)
      }
    }
}

  #return a list with two tibbles
  api_data <- list(api_data = parsed_data, api_metadata = var_metadata, jwmnptst = jwmnptst, tst = tst)

  return(api_data)
}
```

##### Testing: API Query Function - all required values
```{r}
num_vars <- "AGEP, GASP, GRPIP, JWAP, JWDP, JWMNP"
chr_vars <- "FER, HHL, HISPEED, JWTRNS, SCH, SCHL, SEX"
geography <- "state"
key <- "bdb1f6ff2e4982a1da64cd526532aa92dca5581c"
state <- "05"  # Arkansas
year <- 2022

pums_data <- get_PUMS(geography,num_vars,chr_vars,key,year,state)
```
##### Testing: API Query Function default values (`NULL` case for `num_vars` and `chr_vars`)
```{r}
num_vars <- NULL
chr_vars <- NULL
geography <- "state"
key <- "bdb1f6ff2e4982a1da64cd526532aa92dca5581c"
state <- "05"  # Arkansas
year <- 2022

pums_data_default <- get_PUMS(geography,num_vars,chr_vars,key,year,state)
```

##### Testing:  API return
###### First 20 rows 
```{r}
#glimpse the first 20 rows
pums_data$api_data |>
  head(20) |>
  glimpse()
```

###### Testing: Unique factor levels
```{r}
factor_vars <- pums_data$api_data |>
  select(where(is.factor))

factor_vars |> 
  map(unique)
```

##### Testing: Time value mapping
```{r}
num_vars <- "JWAP, JWDP, JWMNP"
# num_vars <- NULL
# chr_vars <- "FER, HHL, HISPEED, JWTRNS, SCH, SCHL, SEX"
chr_vars <- NULL
geography <- "state"
key <- "bdb1f6ff2e4982a1da64cd526532aa92dca5581c"
state <- "05"  # Arkansas
year <- 2022

pums_data <- get_PUMS(geography,num_vars,chr_vars,key,year,state)

time_vars <- pums_data$api_data |>
  select(jwap_t, jwdp_t) |>
  filter(!is.na(jwap_t) & !is.na(jwdp_t)) |>
  head(20) |>
  glimpse()
```

### 7 and 8: Calling a function to query multiple years of API data

```{r}
PUMS_multi_year <- function(geography, num_vars = NULL, chr_vars = NULL, years,
                            key = Sys.getenv("CENSUS_API_KEY"),
                            state = "*", county = NULL, zcta = NULL,
                            survey = "acs1", show_call = FALSE) {
  
  # Empty list to store data for each year
  list_data <- list()
  
  # Looping through each year and calling the get_PUMS function
  for (year in years) {
    year_data <- get_PUMS(geography = geography,
                          num_vars = num_vars,
                          chr_vars = chr_vars,
                          key = key,
                          year = year,
                          state = state,
                          county = county,
                          zcta = zcta,
                          survey = survey,
                          show_call = show_call)
    
    # Extract the 'api_data' part of the year_data list
    year_data <- year_data$api_data
    
    # Ensure year_data is a tibble/data.frame
    if (is.list(year_data)) {
      year_data <- as_tibble(year_data)
    }
    
    # Adding new column to data
    year_data <- year_data |>
      mutate(year = as.character(year))
    
    # Adding data for the year to our list
    list_data[[as.character(year)]] <- year_data
  }
  
  # Combine all yearly data into a tibble
  years_tibble <- bind_rows(list_data)
  
  # Add a "census" class, workflow step 8
  class(years_tibble) <- c("census", class(years_tibble))
  
  return(years_tibble)
}

```

###### Testing Multi-year Query Function - Default Variables

```{r}
# Testing PUMS_multi_year function

# Defining specific years to query
years <- c(2022, 2021)
num_vars <- NULL
chr_vars <- NULL
PUMS_multi_year_test <- PUMS_multi_year(geography, num_vars, chr_vars, years, key, state)
print(unique(PUMS_multi_year_test$year))

```


### 9: Writing a Generic Function for Summarizing
```{r}
summary.census <- function(tibble_class_census,
                           numeric_variables = NULL,
                           categorical_variables = NULL) {

  # Selecting numerical variables
  numeric_vector <- names(select_if(tibble_class_census, is.numeric))

  # Weight vector
  weight_vector <- as.numeric(tibble_class_census$PWGTP)

  # Selecting categorical variables
  categorical_vector <- names(select_if(tibble_class_census, is.character))

  # List to store census summary results
  census_summary_results <- list()

  # Function to obtain weighted mean
  weighted_mean <- function(numeric_vector, weight_vector = NULL) {
    if (!is.null(weight_vector)) {
      return(sum(numeric_vector * weight_vector, na.rm = TRUE) /
               sum(weight_vector, na.rm = TRUE))
    } else {
      return(mean(numeric_vector, na.rm = TRUE))
    }
  }

  # Function to find weighted standard deviation
  weighted_sd <- function(numeric_vector, weight_vector = NULL) {
    sample_mean <- weighted_mean(numeric_vector, weight_vector)
    if (!is.null(weight_vector)) {
      return(sqrt(sum(weight_vector * (numeric_vector^2), na.rm = TRUE) /
                    sum(weight_vector, na.rm = TRUE) - sample_mean^2))
    } else {
      return(sd(numeric_vector, na.rm = TRUE))
    }
  }

  # Numeric summarizer
  for (num_var in numeric_variables) {
    var_data <- as.numeric(tibble_class_census[[num_var]])
    mean_val <- weighted_mean(var_data, weight_vector)
    sd_val <- weighted_sd(var_data, weight_vector)

    # Storing results in list
    census_summary_results[[num_var]] <- list(mean = mean_val, sd = sd_val)
  }

  # Categorical summarizer
  for (cat_var in categorical_variables) {
    counts <- table(tibble_class_census[[cat_var]])
    census_summary_results[[cat_var]] <- counts
  }

  # Plotting a boxplot of numeric variable by categorical variable
  if (!is.null(numeric_variables) && !is.null(categorical_variables)) {
    for (num_var in numeric_variables) {
      for (cat_var in categorical_variables) {
        ggplot(tibble_class_census, aes(x = .data[[cat_var]], y = .data[[num_var]])) +
          geom_boxplot() +
          labs(title = paste("Boxplot of", num_var, "by", cat_var),
               x = cat_var,
               y = num_var) +
          theme_minimal() -> plot

        print(plot)  # Print the plot
      }
    }
  }
  else print("Not plotting")

  return(census_summary_results)
}

# Defining user variables
numeric_variables <- c("PWGTP", "AGEP")
categorical_variables <- c("SEX")

# Testing Summarizer Function
census_summary <- summary.census(tibble_class_census = PUMS_multi_year_test,
                                 numeric_variables = numeric_variables,
                                 categorical_variables = categorical_variables)

# Printing named lists of means, standard deviations, and categorical counts
print(census_summary)


#Defining user variables
# numeric_variables <- c("PWGTP", "AGEP")
# categorical_variables <- c("SEX")
# 
# # Testing Summarizer Function
# census_summary <- summary.census(tibble_class_census = PUMS_multi_year_test,
#                                  numeric_variables = numeric_variables,
#                                  categorical_variables = categorical_variables)
# # Printing named lists of means, standard deviations, and categorical counts
# print(census_summary)

```


###### Testing summary with all vars
```{r}
# Assuming you have PUMS_multi_year_test as your data
census_summary <- summary.census(tibble_class_census = PUMS_multi_year_test,
                                 numeric_variables = num_vars,
                                 categorical_variables = chr_vars)

# Printing named lists of means, standard deviations, and categorical counts
print(census_summary)
```

###### Testing Summarizer Function
```{r}
#Defining user variables
numeric_variables <- c("PWGTP", "AGEP")
categorical_variables <- c("SEX")
# Testing Summarizer Function
census_summary <- summary.census(tibble_class_census = PUMS_multi_year_test,
                                 numeric_variables = numeric_variables,
                                 categorical_variables = categorical_variables)
# Printing named lists of means, standard deviations, and categorical counts
print(census_summary)
```
###### Testing Multi-year Query Function + Summarizer  - All Variables
```{r}
num_vars <- c("AGEP","GASP", "GRPIP", "JWAP", "JWDP", "JWMNP")
# num_vars <- NULL
chr_vars <- c("FER", "HHL", "HISPEED", "JWTRNS", "SCH", "SCHL", "SEX")
# chr_vars <- NULL
geography <- "state"
key <- "bdb1f6ff2e4982a1da64cd526532aa92dca5581c"
state <- "05"  # Arkansas
years <- c(2022, 2021)
PUMS_multi_year_test_all <- PUMS_multi_year(geography, num_vars, chr_vars, years, key, state)

# print(unique(PUMS_multi_year_test_all$year))
```

### 10. Other Analysis - Selected box plots
```{r}

ggplot(PUMS_multi_year_test_all, aes(x = SEX, y = AGEP)) +
  geom_boxplot() +
  ggtitle("Box plot of Age (AGEP) by Sex") +
  theme_minimal()


ggplot(PUMS_multi_year_test_all, aes(x = HISPEED, y = AGEP)) +
  geom_boxplot() +
  ggtitle("Box plot of Broadband Access (HISPEED) by Sex") +
  theme_minimal()

agep_lt30 <- PUMS_multi_year_test_all |>
  filter(AGEP < 30, !is.na(SCH))

agep_lt30$SCH <- recode(agep_lt30$SCH,
                        "N/A (less than 3 years old)" = "N/A (Age < 3)",
                        "Yes, private school or college or home school" = "Private/Home School",
                        "No, has not attended in the last 3 months" = "Not Attending",
                        "Yes, public school or public college" = "Public School")

agep_lt30$SCH <- factor(agep_lt30$SCH)

ggplot(agep_lt30, aes(x = SCH, y = AGEP)) +
  geom_boxplot() +
  ggtitle("Box plot of School Enrollment (SCH) by Age (AGEP) < 30") +
  theme_minimal()

ggplot(agep_lt30, aes(x = SCH , y = AGEP)) +
  geom_boxplot() +
  facet_wrap(~ SEX) +  # Create separate plots for each level of SEX
  theme_minimal() +
  ggtitle("Box plot of Age < 30 vs. School Enrollment by SEX")

```

